Studying measurement invariance in factor analysis models via stochastic processes

### 2011-01-12 MIpaper/par_instable.tex
merged theory revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@18 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

The tests proposed here are generally relevant to situations where the
$p$-dimensional 
data vector $\bm{x}_i, i=1,\cdots,n$ is specified to arise from a model with
density $f(\bm{x}_i | \bm{\theta}_i)$ and log-likelihood 
$\log L(\bm{\theta}_i | \bm{x}_i)$.  
For the factor-analytic situations
considered in this paper, the $\bm{x}_i$ consist of individuals' scale
scores, and 
$f()$ and $\log L()$ arise from the multivariate normal distribution.
The density is defined as:
\begin{equation}
    \label{eq:mvndensity}
    f(\bm{x}_i | \theta_i) = \frac{1}{(2\pi)^{p/2} |
      \bm{\Sigma}(\bm{\theta}_i) |^{1/2}}\exp(-\frac{1}{2}(\bm{x}_i -
    \bm{\mu}(\bm{\theta}_i))^{\prime} \bm{\Sigma}(\bm{\theta}_i)^{-1} (\bm{x}_i -
    \bm{\mu}(\bm{\theta}_i)),
\end{equation}
where ${\bm{\mu}}({\bm{\theta}}_i)$ is the
model-implied mean vector, and ${\bm{\Sigma}}({\bm{\theta}}_i)$ is the
model-implied covariance matrix.
The casewise log-likelihood is defined as:
\begin{equation}
    \label{eq:caselik}
    \log \text{L}({\bm{x}}_i, {\bm{\theta}}_i) = %\sum_{i=1}^{N} 
\log \mid 
\bm{\Sigma}(\bm{\theta}) \mid + %\sum_{i=1}^{N} 
(\bm{x}_i -
\bm{\mu}(\bm{\theta}))^{\prime} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
\bm{\mu}(\bm{\theta})).
\end{equation}
This equation is often summed over $n$ observations, and it is also
often employed for factor analysis and structural
equation models with missing data (e.g., \citeNP{Wot00}).  We use
the likelihood here to facilitate the proposed tests.

Within the general framework outlined above, model estimation
often proceeds by assuming that $\bm{\theta}_i = \bm{\theta}\
\forall\ i$, then choosing $\hat{\bm{\theta}}$ to satisfy the
criterion:
\begin{equation}
    \label{eq:score}
  \sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}}) = 0,    
\end{equation}
 where $\psi(\bm{x}_i, \bm{\theta})$ 
is a score function based on the model.  
For example, in factor analysis models estimated via Maximum
Likelihood, the score function is the first derivative of $\log
\text{L}(\bm{x}_i, \bm{\theta})$ with respect to $\bm{\theta}$.
General M-estimation methods may also be handled within this
framework, though we do not discuss them here.

@@

Measurement invariance specifically implies that factor analysis model
parameters are invariant across individuals.  It can be framed as the
hypothesis:
\begin{equation}
    \label{eq:h0}
    H_0: {\bm \theta}_i = {\bm \theta}_0,\ i=1,\ldots,n,
\end{equation}
signifying that the parameter vector is the same for all individuals
$i$.  
This general hypothesis is very difficult to test because $n$ is large, and
there are many ways by which the $\bm{\theta}_i$ may stray from one
another.  To circumvent this problem, researchers often measure
individuals on an auxiliary variable $V$ and divide them into groups
based on that variable.  
Assuming a continuous $V$, we might divide 
individuals into two groups based on some threshold $\eta$.
Equation~\eqref{eq:h0} then becomes:
\begin{equation}
    \label{eq:h0.cut}
    H_0: \bm{\theta}_{V < \eta} = \bm{\theta}_{V > \eta},
\end{equation}
where, i.e., $\bm{\theta}_{V < \eta}$ reflects the factor analysis
parameters for all individuals with $V < \eta$.

Once we have grouped individuals based on $V$,
factor-analytic assessments
of measurement invariance are well developed.
These assessments are often carried out using nested Multiple Groups
Models (e.g., \citeNP{Jor71,Bol89}) coupled with Likelihood Ratio tests,
though 
Lagrange Multiplier and Wald tests may also be constructed for this
purpose.    While these three tests
are asymptotically equivalent \cite{Sat89}, each is detailed
individually below.

@@

\subsection{Likelihood Ratio Test}

Measurement invariance is achieved
if sets of factor analysis model parameters are equal across
subgroups of individuals 
(\citeNP{Mer93}; there are many types of measurement invariance based
on different sets of model parameters being held constant, but any of these
can be examined via the proposed tests).  If measurement invariance is
not achieved, 
then the parameters assume different values for different
subgroups.  This leads to statistical tests of measurement
invariance via the comparison of two models: a restricted model where
parameters are equal across subgroups, and a ``full'' model
where parameters are allowed to assume different values for each
subgroup.  Statistically, this comparison is often 
accomplished by a likelihood ratio test (also called a
difference test).  Let the full model parameters be denoted 
${\bm \theta} = ({\bm \theta}_1^{\prime}\ {\bm
  \theta}_2^{\prime})^{\prime}$, where we can obtain the restricted model
by placing a set of equality constraints on ${\bm \theta}_1$ described
by the vector-valued function ${\bm
  h}({\bm \theta}_1) = {\bm 0}$.  A $\chi^2$ statistic testing
measurement invariance may then be calculated via:
\begin{equation}
    \label{eq:chisq}
 \chi^2 = -2(n-1)(\log L({\bm X},
\widehat{{\bm \theta}}_c) - \log L({\bm X},\widehat{{\bm \theta}})),
\end{equation}
where ${\bm{X}}$ is an $n \times p$ data matrix, $\widehat{\bm \theta}_c$
contains Maximum Likelihood parameter estimates satisfying the
constraints ${\bm h}({\bm \theta}_1) = {\bm 0}$, and $\widehat{\bm
  \theta}$ contains Maximum Likelihood parameter estimates from the
full model.  The log-likelihood function in the above equation arises
from the multivariate normal distribution and is defined as:
  The test resulting from
Equation~\eqref{eq:chisq} has 
degrees of freedom equal to the difference in number of free
parameters for the two models, and it rejects the hypothesis of measurement
invariance for large values of $\chi^2$.

While the Likelihood Ratio Test is most popular for testing
measurement invariance, Lagrange Multiplier tests and Wald tests
may also be constructed for the same purpose.The Lagrange Multiplier and Wald 
tests, described below, are also advantageous in that they require
estimation of only one 
model (either the full or restricted model), whereas the Likelihood
Ratio Test requires estimation of both models.

\subsection{Lagrange Multiplier Test}
The Lagrange Multiplier test (e.g., \citeNP{Sil59,ChouBent90})
focuses on a restricted model and tests whether the equality
constraints described by ${\bm h}({\bm \theta}_1) = {\bm 0}$ should be
freed.
Estimating the restricted model, we define a vector of Lagrange
multipliers $\widehat{\bm \kappa}$ that satisfy:
\begin{equation}
    \label{eq:lagrange}
    \frac{\partial \log L}{\partial {\bm \theta}} \bigg |_{\widehat{\theta}_c} +
    \frac{\partial {\bm h}}{\partial {\bm \theta}}
    \bigg |_{\widehat{\theta}_c}^{\prime} \widehat{{\bm \kappa}} = {\bm 0}.
\end{equation}
The test itself is then based on the asymptotic distribution of 
$\widehat{{\bm \kappa}}$: if $\widehat{{\bm \kappa}}$ is sufficiently close
to ${\bm 0}$, then the constraints are deemed to hold in the
population.  The resulting test statistic, which is a function of
$\widehat{\bm \kappa}$, follows a $\chi^2$ distribution for large
samples.

\subsection{Wald Test}
In contrast to the Lagrange Multiplier test, the
Wald test (e.g., \citeNP{Wald43})
focuses on the full model and tests the plausibility of the
constraints ${\bm h}({\bm \theta}_1) = {\bm 0}$.
The test is based on the estimates ${\bm h}(\widehat{{\bm \theta}}_1)$
from the full model; if
these are sufficiently close to zero, then the constraints are deemed
to hold in the population.  The resulting test statistic, which is a
function of ${\bm h}(\widehat{{\bm \theta}}_1)$, is also
$\chi^2$-distributed in large 
samples.
Chou and Bentler \citeyear{ChouBent90} compared the 
Wald and Lagrange tests with finite sample sizes via
simulation, and they generally found similar
performance.  They noted, 
however, that the Lagrange test has a tendency toward Type I errors in
some contexts.  They therefore urged caution in the Lagrange test's
use, as it may lead one to discard more constraints than is necessary,
leading to overly-complex models.

\subsection{Extensions to Proposed Tests}
For continuous $V$, the three tests described above require the
threshold $\eta$ to be known.
  This is a
major limitation in psychometrics, where $\eta$ is often unknown or
undefined.
For example, if $V$ represents yearly income, there are many possible
values of 
$\eta$ that could be used to divide individuals into poorer and richer
groups.  The ultimate $\eta$ that we
choose could potentially impact our
conclusions about whether or not a scale is measurement invariant, 
in the same way that dichotomization of continuous variables impacts
general statistical analyses \cite{MacZha02}.

Instead of choosing a specific $\eta$, we may elect to test
individuals across all possible values of $\eta$.  For example, we
could conduct a Likelihood Ratio Test for every possible
dichotomization of $V$, yielding a test statistic for each
dichotomization.  The distribution of the maximum of these test
statistics is derived in Andrews \citeyear{And93}, providing a test of
measurement invariance over all values of $\eta$.  
This is a special case of the family of tests proposed here.  The
tests are designed for situations where
$\eta$ is unknown or where a dichotomization of $V$ is insufficient to
detect measurement invariance violations, and they can be applied both
to individual model parameters and sets of model parameters.

We now describe some general ideas for capturing measurement
invariance violations with respect to a continuous auxiliary variable
$V$.

\section{Stochastic Processes for Measurement Invariance}
The proposed tests of measurement invariance implicitly rely on an
estimated factor analysis model (one that restricts parameters to be
invariant across subgroups).  For the purposes of this paper, we
assume that the model is estimated via Maximum Likelihood.  
The factor analysis log-likelihood may be obtained by summing 
Equation~\eqref{eq:caselik} across individuals.  
Maximization of the likelihood then involves iterative 
algorithms that seek values $\widehat{{\bm{\theta}}}$ to satisfy
Equation~\eqref{eq:score}, where 
$\psi(\bm{x}_i, \theta_j) = \frac{\partial}{\partial \bm{\theta}} \log
\text{L}({\bm{x}}_i, 
{\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}$.
% insert derivatives of casewise evaluated at theta-hat = 0
Such derivatives are taken across individuals, but, using individual
terms of the log-likelihood (i.e., Equation~\eqref{eq:caselik}),
we can also 
evaluate the derivatives with respect to each individual.
If a unique value of $\widehat{{\bm{\theta}}}$ maximized
each individual likelihood, then we would have:
% Equation: \psi(y_i, \theta) = d/d theta casewise_i = 0 \forall i.
\begin{equation}
    \label{eq:casederiv}
    \psi({\bm{x}}_i, \bm{\theta}) = \frac{\partial}{\partial \bm{\theta}} \log
\text{L}({\bm{x}}_i, 
{\bm{\theta}}) \bigg |_{\bm{\theta} = \widehat{\bm{\theta}}} = \bm{0}\ \ \ \forall\ i.
\end{equation}
In practice, no value of $\widehat{{\bm{\theta}}}$ will maximize all
individuals' likelihoods, but, if measurement invariance is satisfied, 
we can expect the $\psi(\cdot, \cdot)$ to fluctuate around zero.  This
is the idea underlying the tests described here.

\subsection{Theory}
Theoretical work in statistics (e.g.,
\citeNP{BroDur75,HjoKon02,Nyb89}) has shown 
that functions of the 
$\psi(\cdot, \cdot)$ follow specific stochastic processes, allowing
us to use them to generally test measurement
invariance.  
We rely on the null hypothesis specified in Equation~\eqref{eq:h0},
which focuses on individuals instead of groups of individuals.
We now review some of the main results
for testing this $H_0$; more
detailed accounts include Hjort and Koning \citeyear{HjoKon02} and
Zeileis and Hornik \citeyear{ZeiHor07}.

% New subsection starts here?

For application to measurement invariance, the most important
theoretical result involves the fact that, under $H_0$, functions of
the $\psi(\cdot, \cdot)$ result in specific stochastic
processes.  One such function
is the {\em{cumulative score process}}, given as:
\begin{equation}
    \label{eq:cumscore}
    W_n(t,\theta_j) = \frac{1}{\sqrt{n}} \sum_{i=1}^{\lfloor nt
      \rfloor} \psi ({\bm{x}}_i, \theta_j),
\end{equation}
where $t$ is a continuous variable in $(1/n,1)$ and $\lfloor nt
\rfloor$ is the floor operator, defined as the integer part of
$nt$.
As the equation shows, the cumulative score process adds subsets of
casewise first derivatives across individuals who have been ordered
based on an auxiliary variable.  Starting at $t=1/n$, only the
first individual's derivative enters into the summation.  Once
$t=2/n$, the first two individuals' derivatives enter into the
summation.  This procedure continues until we reach $t=n$, at which
point all individuals' first derivatives have entered into the
summation.  There are $n$ cumulative sums obtained by sequentially
inserting individuals' derivatives into the equation, and these are
calculated 
for each of the $p$ model parameters.  This results
in $n\times p$
empirical cumulative scores.

Elements of the summation in Equation~\eqref{eq:cumscore}
can be positive or 
negative, so that the cumulative scores do not necessarily increase
with $t$.  In fact, if measurement invariance (i.e., $H_0$) holds,
then the cumulative scores should 
follow a stochastic process that fluctuates around zero.  
Zeileis and Hornik \citeyear{ZeiHor07} showed specifically that, under
$H_0$:
\begin{equation}
    \label{eq:browbridge}
{\bm{B}}(\cdot, \widehat{{\bm{\theta}}}) = {\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}})
\overset{d}{\rightarrow} {\bm W}^{0}(\cdot), % put d over rightarrow
\end{equation}
where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed information
matrix (e.g., of the factor analysis model), ${\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}})$ reflects the $p$-dimensional
cumulative sums (one for each parameter), ${\bm W}^{0}(\cdot)$ is a
$p$-dimensional Brownian bridge, and 
$\overset{d}{\rightarrow}$ reflects convergence in distribution.
In words, there are $p$ cumulative score processes, one for each
model parameter.  This collection of processes follows a
multidimensional Brownian bridge as derivatives accumulate in the 
summation from individual 1 to individual $n$.
Multiplication by ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}$ 
``decorrelates'' the $p$
cumulative score processes, such that each univariate process (i.e., each
process for a single model parameter) is unrelated to all other
processes (and is asymptotically independent of the other processes).

The cumulative scores resulting from Equation~\eqref{eq:browbridge}
can be arranged in an $n \times p$ matrix called an {\em{empirical
    cumulative score process}}, which we will denote as
${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$.
Each column of this matrix follows a univariate Brownian bridge and is
based on a single factor analysis parameter. 
To carry out a test of $H_0$, we need functions that collapse across
rows (individuals) and/or columns (parameters) of the matrix.  
These
functions then lead to scalar test statistics and p-values.
The specific choice of function depends on the goals of the analyst:
the row-collapsing functions are useful for isolating subsets of
parameters that violate measurement invariance, and the
column-collapsing functions are useful for isolating subgroups of
individuals who violate measurement invariance across all parameters.
Some functions result in test
statistics whose critical values can be derived analytically, while
others require critical values to be derived via the
simulation of a Brownian bridge.

>>>>

\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

%% for internal use
\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}

@@

The methods proposed here are generally relevant to situations where the
$p$-dimensional random variable $X$ with associated observations $\bm{x}_i, i=1,\cdots,n$
is specified to arise from a model with density $f(\bm{x}_i; \bm{\theta})$ and
associated joint log-likelihood
\begin{equation} \label{eq:loglik}
  \ell(\bm{\theta}; \bm{x}_1, \dots, \bm{x}_n) ~=~
    \sum_{i = 1}^n \ell(\bm{\theta}; \bm{x}_i) ~=~
    \sum_{i = 1}^n  \log f(\bm{x}_i; \bm{\theta}),
\end{equation}
where ${\bm \theta}$ is some $k$-dimensional parameter vector that characterizes
the distribution. 
The methods are applicable under very general conditions, essentially
whenever standard assumptions for maximum likelihood inference hold (for more
details see below). For empirical applications, we employ
a specific factor analysis model where the data follow a multivariate normal
distribution:
\begin{eqnarray}
    \label{eq:mvndensity}
    f(\bm{x}_i; \bm{\theta}) & = & \frac{1}{(2\pi)^{p/2} |
      \bm{\Sigma}(\bm{\theta}) |^{1/2}} ~ \exp \left\{ -\frac{1}{2}(\bm{x}_i -
    \bm{\mu}(\bm{\theta}))^{\top} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
    \bm{\mu}(\bm{\theta}) \right\}, \\
    \label{eq:caselik}
    \ell(\bm{\theta}; \bm{x}_i) & = & -\frac{1}{2} \left\{
      (\bm{x}_i - \bm{\mu}(\bm{\theta}))^{\top} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i - \bm{\mu}(\bm{\theta}))
      ~+~ \log | \bm{\Sigma}(\bm{\theta}) | ~+~ p \log(2 \pi) \right\}.
\end{eqnarray}
with model-implied mean vector ${\bm{\mu}}({\bm{\theta}})$ and
covariance matrix ${\bm{\Sigma}}({\bm{\theta}})$. As pointed out above,
the assumptions for the tests introduced below do not require this specific
form of the likelihood but it is presented for illustration due to
its importance in practice.  Many expositions of factor analysis
utilize the likelihood for the sample covariance matrix, which is
based on a Wishart distribution when $\bm{x}_i$ are assumed to be
multivariate normal.
However, the techniques presented below require the casewise contributions
to the likelihood; this situation is also generally
encountered in structural equation models with missing data (e.g.,
\citeNP{Wot00}).

Within the general framework outlined above, and under the usual
regularity conditions, the model parameters $\bm{\theta}$ can
be estimated by maximum likelihood (ML), i.e.,
\begin{equation} \label{eq:ml}
  \hat{\bm{\theta}} ~=~ \argmax_{\bm{\theta}} \ell(\bm{\theta}; x_1, \dots, x_n),
\end{equation}
or equivalently by solving the first order conditions
\begin{equation}
    \label{eq:ml1}
  \sum_{i=1}^{n} {\bm s}(\hat{\bm{\theta}}; \bm{x}_i) ~=~ 0,    
\end{equation}
where 
\begin{equation}
  \label{eq:score}
  {\bm s}({\bm \theta}; x_i) ~=~ \left(
    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_1},
    \dots,
    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_k}
  \right)^\top,
\end{equation}
is the score function of the model, i.e., the partial derivative of the casewise
likelihood contributions w.r.t.\ the parameters $\bm{\theta}$.

% \fixme{I have deferred individual-specific parameters for ease of exposition.
% My feeling is that it should be briefly mentioned at the end of this
% section and then connected with measurement invariance in the next section.}

One central assumption -- sometimes made implicitly -- is that
the same model parameters $\bm{\theta}$ hold for all individuals $i = 1, \dots, n$.
If this is not satisfied, the estimates $\hat{\bm{\theta}}$ are typically
not meaningful and cannot easily be interpreted.
One potential source of deviation from this assumption is lack of
measurement invariance, investigated in the following section.

@@

In general terms, a set of scales is defined to be measurement
invariant with respect to an auxiliary variable $V$ if:
\begin{equation}
    \label{eq:midef}
      f({\bf X} | T, V, \dots) = f({\bf X} | T, \dots),
\end{equation}
where ${\bf X}$ is a data matrix, $T$ is the latent
variable that the scales purport to measure, and $f$ is the model's
distributional form.  In the parametric framework
adopted here, this means that the parameter vector ${\bm \theta}$ (or
some subset of ${\bm \theta}$; see \citeNP{Mer93}) is equal
across subgroups of individuals and thus does not vary
with any variable $V$.

To frame this as a formal hypothesis, we assume that -- in principle --
Model~(\ref{eq:loglik}) holds for all individuals but with a
potentially individual-specific 
parameter vector ${\bm \theta}_i$. The null hypothesis of measurement invariance
is then equivalent to the null hypothesis of parameter constancy
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
which should be tested against the alternative that the parameters
are some nonconstant function ${\bm \theta}(\cdot)$ of the variable $V$ with observations
$v_1, \dots, v_n$, i.e.,
\begin{equation}
    \label{eq:h1}
    H_1:~ {\bm \theta}_i = {\bm \theta}(v_i),\quad (i=1,\ldots,n).
\end{equation}
where the pattern ${\bm \theta}(V)$ of deviation from measurement invariance is
typically not known (exactly) in practice. If it were (see below for
some concrete examples), then standard inference methods (such as likelihood
ratio, Wald, or Lagrange multiplier tests) could be employed. However,
if the pattern is unknown, it is difficult to develop a single test
that is well-suited for all conceivable patterns. But it is possible to
derive a family of tests so that representatives from this family are well-suited
for a wide range of possible patterns. One pattern of particular interest
involves $V$ dividing the individuals into two subgroups with different
parameter vectors
\begin{equation}
  \label{eq:h1*}
  H_1^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where ${\bm \theta}^{(A)} \neq {\bm \theta}^{(B)}$. This could pertain to 
two different age groups, income groups, genders, etc.

Note that even when adopting $H_1^*$ as the relevant alternative, the pattern ${\bm \theta}(V)$
is not completely specified unless the cutpoint $\nu$ is known in
advance. 
In this situation, all individuals can be grouped based on
$V$, and we can apply standard
theory: nested multiple group
models (e.g., \citeNP{Jor71,Bol89}) coupled with likelihood ratio (LR)
tests are most common,
though Lagrange multiplier (LM) and Wald tests may also be constructed for this
purpose.\footnote{All three of these approaches are asymptotically
  equivalent \cite{Sat89}.}  However, if $\nu$ is unknown (as is often
the case for continuous $V$), then the standard theory does not apply
and we can instead rely on the tests proposed in this paper.

In the following section, we describe the standard approaches to
testing measurement invariance with $\nu$ known.  We then contrast
these approaches with the tests proposed in this paper.
We assume throughout that the observations $i = 1, \dots, n$ are
ordered with respect to the random variable $V$ of interest such that
$v_1 \le v_2 \le \dots \le v_n$.

@@

\subsection{Likelihood Ratio, Wald, and Lagrange Multiplier Test for Fixed Subgroups}

To employ the LR test for assessing measurement invariance, the data
are typically separated into a certain number of given subgroups where
the parameters are estimated separately (for ease of
  exposition, we assume no parameters are restricted to be equal
  across subgroups). Then, the sum of maximized likelihoods 
from the subgroups are compared with the original maximized full-sample likelihood
in a $\chi^2$~test. More specifically, for the special case of two subgroups,
the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
for the given threshold $\nu$ is then
\begin{equation} \label{eq:lr}
  \mathit{LR}(\nu) ~=~ -2 \left[
         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
    \right],
\end{equation}
which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
of parameters in~${\bm \theta}$.
\readme{You previously modified the above section so that $\theta$
  consists only of parameters that are ``doubled'' under the
  alternative.  This appears to not allow for situations where
  parameters are restricted to be equal across subgroups.  I am ok
  with this, but most applications of LR tests with these models have
  some parameters restricted to be equal across subgroups.  So I've
  added some detail in parentheses at the beginning of the above
  paragraph.}

Analogously to the LR test, the Wald test and LM test (also known as score test)
can be employed to test the null hypothesis $H_1^*$ for a fixed threshold $\nu$.
For the Wald test, the idea is to compute the Wald statistic $W(\nu)$ as a quadratic
form in $\hat {\bm \theta}^{(A)} - \hat {\bm \theta}^{(B)}$, utilizing its estimated
covariance matrix for standardization. For the LM test, the LM statistic
$\mathit{LM}(\nu)$ is a quadratic form in ${\bm s}(\hat {\bm \theta}; x_1, \dots, x_m)$
and ${\bm s}(\hat {\bm \theta}; x_{m+1}, \dots, x_n)$. Thus, the three tests all assess
differences that should be zero under $H_0$: for the LR test the difference of maximized
likelihoods; for the Wald test, the difference of parameter estimates; and for the
LM test, the differences of likelihood scores from zero. In the LR case, the parameters
have to be estimated both under null hypothesis and alternative while in the Wald case
only the estimates under the alternative are required and in the LM case only under the
null hypothesis.

\subsection{Extensions for Unknown Subgroups}

For assessing measurement invariance in psychometric models, the major limitation of
the three tests is that the potential subgroups have to be known in advance. 
Even if the variable $V$ w.r.t.\ which the violation of invariance occurs is
known, the threshold $\nu$ from (\ref{eq:h1*}) is often unknown in practice.
For example, if $V$ represents yearly income, there are many possible
values of $\nu$ that could be used to divide individuals into poorer and richer
groups.  The ultimate $\nu$ that we
choose could potentially impact our
conclusions about whether or not a scale is measurement invariant, 
in the same way that dichotomization of continuous variables impacts
general psychometric analyses \cite{MacZha02}.

Instead of choosing a specific $\nu$, a natural idea is to compute $\mathit{LR}(\nu)$
for each possible value in some interval $[\underline{\nu}, \overline{\nu}]$ and
reject if their maximum
\begin{equation} \label{eq:maxlr}
  \max_{\nu \in [\underline{\nu}, \overline{\nu}]} \mathit{LR}(\nu)
\end{equation}
becomes large. Note that this corresponds to maximizing the likelihood w.r.t.\
an additional parameter, namely $\nu$. Hence, the asymptotic distribution of the
maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. However, Andrews \citeyear{And93}
showed that the asymptotic distribution is in fact tractable, but nonstandard.
Specifically, the asymptotic distribution of the maximum is the maximum of a
certain tied-down Bessel process whose specifics also depend on the minimal
and maximal thresholds $\underline{\nu}$ and $\overline{\nu}$, respectively.
See below for more details.

Analogously, one can consider $\max W(\nu)$ and
$\max \mathit{LM}(\nu)$, respectively, which both have the same asymptotic properties
as $\max \mathit{LR}(\nu)$ and are asymptotically equivalent \cite{And93}.
From a computational perspective, the latter test is particularly convenient
because it requires just a single set of estimated parameters $\hat {\bm \theta}$
which is employed for all thresholds $\nu$ in $[\underline{\nu}, \overline{\nu}]$
while the other two techniques require reestimation of the subgroup model for each
$\nu$.

So far, the discussion focused on the alternative $H_1^*$: The maximum LR, Wald, and LM
tests are designed for a situation where there is a single threshold at which the full
parameter vector ${\bm \theta}$ changes. While this is plausible and intuitive
in many applications, it would also be desirable to obtain tests that direct their power
against other types of alternatives, i.e., against $H_1$ with other patterns ${\bm \theta}(V)$.
For example, there might be multiple thresholds at which the parameters change or parameters
fluctuate randomly. Alternatively, only one (or just a few) of the parameters in
the vector ${\bm \theta}$ change while the remaining parameters are
constant (a common occurrence in psychometric models).
To address such situations in a unified way, the next section contains 
a general framework for testing measurement invariance along a
(continuous) variable $V$ that
includes the maximum LM test as a special case.

\section{Stochastic Processes for Measurement Invariance}

As discussed above, factor analysis models are typically
estimated by fitting the model to all $i = 1, \dots, n$ individuals, assuming
that the parameter vector ${\bm \theta}$ is constant across individuals.
Having estimated the parameters $\hat {\bm \theta}$, the goal is to check that
there are no subgroups of individuals that fail to conform with the model (for any
of the parameters). Hence, some measure of model deviation or residual is required
that captures the lack of fit for the $i$-th individual at the $j$-th parameter
($i = 1, \dots, n$, $j = 1, \dots, k$). A natural measure -- that employs the ideas
of the LM test -- is ${\bm s}(\hat {\bm \theta}; x_i)_j$: the $j$-th component of the
contribution of the $i$-th observation to the score function. By construction the sum
of the score contributions over all individuals is zero for each component, see
(\ref{eq:ml1}). Moreover, if there are no systematic deviations the score contributions
should fluctuate randomly around zero while they should be shifted away from zero for
subgroups where the model does not fit.

Therefore, to employ this quantity for tests of measurement invariance against alternatives
of type (\ref{eq:h1}), we need to overcome two obstacles: (1)~make use of the ordering of
the observations w.r.t.\ $V$ because we want to test for changes ``along'' $V$; (2)~account
for potential correlations between the $k$~components of the parameters to be able to detect
which parameter(s) change (if any).

\subsection{Theory}

The test problem of the null hypothesis (\ref{eq:h0}) against the alternatives
(\ref{eq:h1}) and (\ref{eq:h1*}), respectively, has been studied extensively
in the statistics and econometrics literature under the label ``structural change
tests'' (see e.g., \citeNP{BroDur75,And93}) where the focus of interest is the
detection of parameter instabilities of time series models ``along'' time.
Specifically, it has been shown (e.g., \citeNP{Nyb89,Han92,HjoKon02,ZeiHor07})
that cumulative sums of the empirical scores follow specific stochastic processes, allowing
us to use them to generally test measurement invariance. Here, we review
some of the main results from that literature and adapt it to the specific challenges
of factor analysis models. More detailed accounts of the underlying structural
change methods include Hjort and Koning \citeyear{HjoKon02} and
Zeileis and Hornik \citeyear{ZeiHor07}.

For application to measurement invariance, the most important
theoretical result involves the fact that, under $H_0$, the
\emph{cumulative score process} converges to a specific asymptotic
process. The $k$-dimensional cumulative score process is defined as
\begin{equation} \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n t \rfloor} {\bm s}(\hat {\bm \theta}; x_i)
  \qquad (0 \le t \le 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of $nt$ and
$\hat I$ is some consistent estimate of the covariance matrix of the scores,
e.g., their outer product or the observed information matrix.
As the equation shows, the cumulative score process adds subsets of
casewise score contributions across individuals along the ordering
w.r.t.\ the variable $V$ of interest. At $t=1/n$, only the
first individual's contribution enters into the summation; at
$t=2/n$, the first two individuals' contributions etc.\ until $t = n/n$
where all contributions enter the sum. Thus, due to (\ref{eq:ml1}),
the cumulative score process always equals zero at $t = 0$ and
returns to zero at $t = 1$. Furthermore, multiplication by
$\hat {\bm I}^{-1/2}$ ``decorrelates'' the $k$ cumulative score
processes, such that each univariate process (i.e., each
process for a single model parameter) is unrelated to (and asymptotically independent of) all other
processes. Therefore, this cumulative process ${\bm B}(t; \hat {\bm \theta})$
accomplishes the challenges discussed at the beginning of this section:
it makes use of the ordering of the observations by taking cumulative sums
and it decorrelates the contributions of the $k$ different parameters.

Inference can then be based on an extension of the usual central limit
theorem. Under the assumption of independence of individuals (implicit
already in Equation~\ref{eq:loglik}) and under
the usual ML regularity conditions (assuring asymptotic normality
of $\hat {\bm \theta}$), Hjort \& Koning \citeyear{HjoKon02} show
that 
\begin{equation} \label{eq:fclt}
  {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot),
\end{equation}
where $\overset{d}{\rightarrow}$ denotes convergence in distribution
and ${\bm B}^{0}(\cdot)$ is a $k$-dimensional Brownian bridge.
In words, there are $k$ cumulative score processes, one for each
model parameter.  This collection of processes follows a
multidimensional Brownian bridge as score contributions accumulate in the 
summation from individual 1 (with lowest value of $V$) to individual $n$
(with highest value of $V$).

The empirical cumulative score process from (\ref{eq:cumscore}) can also be viewed
as an $n \times k$ matrix with elements ${\bm B}(i/n; \hat {\bm \theta})_j$ that
we also denote ${\bm B}(\hat {\bm \theta})_{ij}$ below for brevity.
Each column of this matrix converges to a univariate Brownian bridge and pertains
to a single factor analysis parameter. To carry out a test of $H_0$, this process/matrix
needs to be aggregated to a scalar test statistic by collapsing across
rows (individuals) and columns (parameters) of the matrix. The asymptotic distribution
of this test statistic is then easily obtained by applying the same
aggregation to the asymptotic process ${\bm B}^0$ \cite{HjoKon02,ZeiHor07},
so that corresponding $p$-values can be derived.
